Rotation Period Evolution in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking

### 2019-04-25 Paper/sync.tex
Updates from Overleaf

We examine how tides, stellar evolution, and magnetic braking shape the rotation period (P$_{rot}$) evolution of low-mass stellar binaries up to orbital periods (P$_{orb}$) of 100 d across a wide range tidal dissipation parameters using two common equilibrium tidal models. We find that many binaries with P$_{orb} \lsim 20$ d tidally-lock, and most with $P_{orb} \lsim 4$ d tidally-lock into synchronous rotation on circularized orbits. At short P$_{orb}$, tidal torques produce a population of fast rotators that single-star only models of magnetic braking fail to produce.  In many cases, we show that the competition between magnetic braking and tides produces a population of subsynchronous rotators that persists for Gyrs, even in short P$_{orb}$ binaries, qualitatively reproducing the subsynchronous eclipsing binaries (EBs) discovered in the \kepler field by \citet{Lurie2017}. Both equilibrium tidal models predict that binaries can tidally-interact out to P$_{orb} \approx 80$ d, while the \xxx{Constant Phase Lag tidal} model predicts that binaries can tidally-lock out to P$_{orb} \approx 100$ d. Tidal torques often force the P$_{rot}$ evolution of stellar binaries to depart from the long-term magnetic braking-driven spin down experienced by single stars, revealing that P$_{rot}$ is not be a valid proxy for age in all cases, i.e. gyrochronology methods can fail unless one accounts for binarity. We suggest that accurate determinations of orbital eccentricties and P$_{rot}$ can be used to discriminate between which equilibrium tidal models best describes tidal interactions in low-mass binary stars.

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../Plots/lockedCPL.pdf}
   \caption{\xxx{Updated Figure} Rotation state for tidally-locked (blue, P$_{rot} = $ P$_{eq}$), interacting (green, P$_{rot}$ within $10\%$ of P$_{eq}$ and not locked), and not locked (orange, remainder of binaries) stellar binaries. Left: P$_{rot}$ as a function of stellar mass and age according to our CPL simulations integrated to system ages uniformly sampled over $1-7$ Gyr. Right: Marginalized P$_{rot}$ distribution for each case.}%
    \label{fig:lockedCPL}%
\end{figure}

@@

We compare the P$_{rot}$ distribution for tidally-interacting stellar binaries from our CPL and CTL simulations with the P$_{rot}$ distribution of single stars subject to stellar evolution and magnetic braking to gauge the impact of tidal torques on driving P$_{rot}$ distributions away from that of single stars.  We simulate 10,000 single star systems according to the evolution described in $\S$~\ref{sec:methods:stellar} with initial conditions sampled from the same mass and P$_{rot}$ distributions used for the binary simulations described $\S$~\ref{sec:methods}. In Fig.~\ref{fig:protDist}, we display P$_{rot}$ as a function of mass and age for binaries simulated using both the CPL and CTL model and for single stars.

@@

\xxx{There is a clear age-dependence of P$_{rot}$ seen in the single star population with older stars rotating more slowly, a trend that is borne out in nature and is the critical assumption of gyrochronology methods that link P$_{rot}$ to stellar ages via the magnetic braking-driven long-term spin down of low-mass stars \citep[e.g.][]{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010,Meibom2015}.
In stark contrast, both tidal models predict that age does not always strongly correlate with P$_{rot}$ as binaries at a given P$_{rot}$ can assume a range of ages, especially at short and long P$_{rot}$, due to the influence of tidal torques. In binaries, tidal torques tend to drive the P$_{rot}$ evolution away from that of single stars and towards P$_{eq}$, either maintaining rapid rotation in tidally-locked short P$_{orb}$ systems, or working with magnetic braking to slow P$_{rot}$ beyond that of single stars of the same age.}

\xxx{We quantify the impact of binarity on gyrochronology in Fig.~\ref{fig:gyro} by computing the percent difference between the mean ages of single and binary stars, for both tidal models, in mass and P$_{rot}$ bins for the populations depicted in Fig.~\ref{fig:protDist}. This quantity, referred to here as the ``Relative Age Error", represents the systematic error incurred by assigning a binary star the age expected for single stars at a given mass and P$_{rot}$. For this comparison, we select the subset of tidally-locked and tidally-interacting binaries as classified in Fig.~\ref{fig:lockedCPL} and Fig.~\ref{fig:lockedCTL}, accounting for $31\%$ and $24\%$ of CPL and CTL binaries, respectively, as the remaining binaries weakly tidally-interact, and hence their P$_{rot}$ evolution will predominately follow that of single stars.}

\begin{figure*}[t]
	\includegraphics[width=\textwidth]{../Plots/gyro.pdf}
   \caption{\xxx{New Figure: Relative age error between single and binary stars for both the CPL (left) and CTL (right) tidal models. The error is computed as the percent difference between the mean ages of single and binary stars in mass and P$_{rot}$ bins.}}%
    \label{fig:gyro}%
\end{figure*}

\xxx{Gyrochronology ages, when applied to tidally-interacting binaries systematically underestimate the ages, with the age error increasing with decreasing P$_{rot}$ and increasing with primary star mass where tides are stronger. For binaries with P$_{rot} \lsim 30 d$, gyrochronology ages are underestimated by $50\%$, with this error growing to $300\%$ for P$_{rot} \approx 10$ d. For slow rotators with P$_{rot} \gsim 40$ d, gyrochronology ages are slightly overestimated by up to $25\%$ for near solar-mass stars. Single star-only models fail to produce near-solar mass stars with P$_{rot} \lsim 10$ d nor P$_{rot} \gsim 50$ as only tidally-interacting binaries at short and long P$_{orb}$, respectively, can produce such rotation states, so we can not compare binary star and single star ages there.}

\xxx{... This effect is most pronounced at short P$_{orb}$, and hence P$_{rot}$, where tidal torques in binaries are strongest, leading to relative age errors of up to $300\%$ for both tidal models relative to single stars. This effect would be more pronounced for the most rapidly-rotating stars, e.g. tidally-interacting near-solar mass binaries with P$_{rot} \lsim 10$ d, however our single star model and age prior cannot produce such rapid rotators as only those with ages $\lsim 1$ Gyr could have P$_{rot} \lsim 10$ d. In Fig.~\ref{fig:protAgeHist}, we further examine the age distributions of rapid rotators.}

\xxx{The age distribution of tidally-interacting binaries with P$_{rot} < 20 $ d is inconsistent with that of single stars. In this range, the median ages and $68\%$ interval are $2.3^{+2.9}_{-0.9}$ Gyr and $2.4^{+3.0}_{-1.1}$ Gyr according to the CPL and CTL models, respectively, compared to the much younger single stars with ages of $1.6^{+0.8}_{-0.4}$ Gyr. We highlight this dichotomy in Fig.~\ref{fig:protAgeHist} by plotting a histogram of system ages from Fig.~\ref{fig:protDist} for single or primary stars in binaries with P$_{rot} < 20$ d.}

@@

Tidal torques pose a fundamental problem for inferring ages of stars via \xxx{gyrochronology}. Regardless of the choice of equilibrium tidal model or magnetic braking model, stellar binaries readily tidally-lock, or at least strongly tidally-interact, across a wide range of P$_{orb}$ and primary star masses, decoupling P$_{rot}$ from age. For example, if one observed a rapidly rotating star with P$_{rot} \lsim 20$ d, gyrochronology models would predict ages \xxx{$\lsim 1.6$} Gyr. If a rapidly-rotating star is actually an unresolved binary, it would likely be tidally-locked, decoupling P$_{rot}$ from age, causing the predictions of gyrochronology models to fail as its age could be at least several Gyrs, e.g. Fig.~\ref{fig:protAgeHist}. This effect is most likely to manifest in rapid rotators (P$_{rot} < 20$ d), but persists across all P$_{rot}$ up to 100 d, producing a significant contaminating signal, \xxx{e.g. Fig.~\ref{fig:protDist} and Fig.~\ref{fig:gyro}}.

In general, it is difficult to accurately determine if a source is single star or a stellar binary via longterm photometric monitoring as only a small fraction of stars in binaries will occult one another. Observations of the binarity of field stars by \citet{Raghavan2010} and \citet{Duchene2013} indicate that roughly half of stars are in stellar binaries, with $10\%$ of these binaries having P$_{orb} \lsim 100$ d, suggesting that unless one accounts for binarity, stellar binaries will produce a non-negligible contaminating signal in any study of stellar rotation periods and any ages inferred via \xxx{gyrochronology} are potentially subject to systematic errors.  Moreover, this problem could be more significant as \citet{Simonian2018} found that most rapid rotators with P$_{rot} \leq 7.5$ d in the \kepler field \xxx{are consistent with tidally-synchronized photometric binaries, suggesting that binary contamination in P$_{rot}$ studies could be widespread}. We caution that any application of \xxx{gyrochronology} methods to predict ages for stars, especially those with P$_{rot} \lsim 20$ d, should rule out or account for stellar binarity, or otherwise risk deriving systematically incorrect ages.

@@

The torques due to tides and magnetic braking balance when $ \frac{dJ}{dt}|_{tides} +  \frac{dJ}{dt}|_{MB} = 0$, 
\begin{equation} \label{eqn:appendixBalance}
    \frac{C_{tides} k_2 \tau}{\mathrm{P}_{orb}^5}\left( 1 - \frac{\mathrm{P}_{orb}}{\mathrm{P}_{rot}} \right) -C_{MB} \left( \frac{P_{rot,\odot}}{P_{rot}} \right)^3 = 0.
\end{equation}
By specifying P$_{orb}$ and the product, $k_2 \tau$, we can numerically solve Eqn.~(\ref{eqn:appendixBalance}) for the P$_{rot}$ at which torques due to magnetic braking and tides balance, often producing subsynchronous rotation as seen in Fig.~\ref{fig:analyticBalance} and our simulations in $\S$~\ref{sec:eq}.

>>>>

We examine how tides, stellar evolution, and magnetic braking shape the rotation period (P$_{rot}$) evolution of low-mass stellar binaries up to orbital periods (P$_{orb}$) of 100 d across a wide range tidal dissipation parameters using two common equilibrium tidal models. We find that many binaries with P$_{orb} \lsim 20$ d tidally-lock, and most with $P_{orb} \lsim 4$ d tidally-lock into synchronous rotation on circularized orbits. At short P$_{orb}$, tidal torques produce a population of fast rotators that single-star only models of magnetic braking fail to produce.  In many cases, we show that the competition between magnetic braking and tides produces a population of subsynchronous rotators that persists for Gyrs, even in short P$_{orb}$ binaries, qualitatively reproducing the subsynchronous eclipsing binaries (EBs) discovered in the \kepler field by \citet{Lurie2017}. Both equilibrium tidal models predict that binaries can tidally-interact out to P$_{orb} \approx 80$ d, while the \xxx{Constant Phase Lag tidal} model predicts that binaries can tidally-lock out to P$_{orb} \approx 100$ d. Tidal torques often force the P$_{rot}$ evolution of stellar binaries to depart from the long-term magnetic braking-driven spin down experienced by single stars, revealing that P$_{rot}$ is not be a valid proxy for age in all cases, i.e. gyrochronology \xxx{ages can err by up to $300\%$} unless one accounts for binarity. We suggest that accurate determinations of orbital eccentricties and P$_{rot}$ can be used to discriminate between which equilibrium tidal models best describes tidal interactions in low-mass binary stars.

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../Plots/lockedCPL.pdf}
   \caption{\xxx{Updated Figure} Rotation state for tidally-locked (blue, P$_{rot} = $ P$_{eq}$), interacting (green, P$_{rot}$ within $10\%$ of P$_{eq}$ and not locked), and not locked (orange, remainder of binaries) stellar binaries. Left: P$_{rot}$ as a function of stellar mass and age according to our CPL simulations integrated to system ages uniformly sampled over $1-7$ Gyr. Right: Marginalized P$_{rot}$ distribution for each case. \xxx{Top: Marginalized mass distributions.}}%
    \label{fig:lockedCPL}%
\end{figure}

@@

We compare the \xxx{P$_{rot}$ and age distributions of tidally-interacting stellar binaries from our CPL and CTL simulations with that of single stars to gauge the impact of tidal torques on driving P$_{rot}$ distributions away from that of single stars and what implications that may have for estimating stellar ages using gyrochronology}.  We simulate 10,000 single star systems according to the evolution described in $\S$~\ref{sec:methods:stellar} with initial conditions sampled from the same mass and P$_{rot}$ distributions used for the binary simulations described $\S$~\ref{sec:methods}. In Fig.~\ref{fig:protDist}, we display P$_{rot}$ as a function of mass and age for binaries simulated using both the CPL and CTL model and for single stars.

@@

\xxx{In binaries, tidal torques tend to drive the P$_{rot}$ evolution away from that of single stars and towards P$_{eq}$, either maintaining rapid rotation in tidally-locked short P$_{orb}$ systems, or working with magnetic braking to slow P$_{rot}$ beyond that of single stars of the same age. The impact of tidal torques on the binary P$_{rot}$ distribution is clear: strong tidal torques in short P$_{orb}$ binaries produce a substantial population of rapid rotators with P$_{rot} \lsim 20$ d, with the influence of tides extending to longer P$_{orb}$ systems as demonstrated in Fig.~\ref{fig:lockedCPL} and Fig.~\ref{fig:lockedCTL}. Except for stars with ages $\lsim 1$ Gyr, or young late M-dwarfs who are either contracting along the pre-main sequence or have just reached the main sequence, our single star simulations fail to produce a population of rapid rotators. This theoretical result is consistent with \citet{Simonian2018} who find that the population of \kepler stars with P$_{rot} < 7.5$ d is likely dominated by tidally-interacting binaries.}

\xxx{In the single star population, there is a clear monotonic age-dependence of P$_{rot}$, with older stars rotating more slowly, a trend that is borne out in nature and is the critical assumption of gyrochronology methods that link P$_{rot}$ to stellar ages via the magnetic braking-driven long-term spin down of low-mass stars \citep[e.g.][]{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010,Meibom2015}. This trend is a generic outcome of magnetic braking and is not specific to our choice of magnetic braking model. In stark contrast, both tidal models predict that age does not always strongly correlate with P$_{rot}$ as tidally-interacting binaries, at a given primary star mass and P$_{rot}$, can assume a wide range of ages, especially for P$_{rot} \lsim 20$ d.}

\xxx{We quantify the impact of binarity on gyrochronology age estimates in Fig.~\ref{fig:gyro} by computing the percent difference between the mean ages of single and binary stars, for both tidal models, in mass and P$_{rot}$ bins for the populations depicted in Fig.~\ref{fig:protDist}. This quantity, referred to here as the ``Relative Age Error", represents the systematic error incurred by assigning a tidally-interacting binary star the age expected for single stars at a given mass and P$_{rot}$. For this comparison, we select the subset of tidally-locked and tidally-interacting binaries as classified in Fig.~\ref{fig:lockedCPL} and Fig.~\ref{fig:lockedCTL}, accounting for $31\%$ and $24\%$ of CPL and CTL binaries, respectively.}

\begin{figure*}[t]
	\includegraphics[width=\textwidth]{../Plots/gyro.pdf}
   \caption{\xxx{New Figure: Relative age error between single and binary stars for both the CPL (left) and CTL (right) tidal models. The error is computed as the percent difference between the mean ages of single and tidally-interacting binary stars in mass and P$_{rot}$ bins.}}%
    \label{fig:gyro}%
\end{figure*}

\xxx{For most values of P$_{rot}$, gyrochronology methods systematically underestimate the ages of tidally-interacting binaries, with the relative age error increasing with decreasing P$_{rot}$ and increasing with primary star mass, where tides are stronger. For binaries with P$_{rot} \lsim 30$ d, gyrochronology ages are underestimated by $50\%$, with this error growing to $300\%$ for P$_{rot} \approx 10$ d. For slow rotators with P$_{rot} \gsim 40$ d, gyrochronology ages are slightly overestimated by up to $25\%$, with the largest errors occurring for near solar-mass primary stars where tides and magnetic braking combine to spin down binary stars.  The relative age errors would be more pronounced for the most rapidly-rotating stars, e.g. tidally-interacting near-solar mass binaries with P$_{rot} \lsim 10$ d, however, our single star-only models fail to produce such rotation states.}

\xxx{The age distribution of binaries with P$_{rot} < 20$ d is inconsistent with that of single stars. In this range, the median ages and $68\%$ interval are $2.3^{+2.9}_{-0.9}$ Gyr and $2.4^{+3.0}_{-1.1}$ Gyr according to the CPL and CTL models, respectively, compared to the much younger single stars with ages of $1.6^{+0.8}_{-0.4}$ Gyr. We highlight this dichotomy in Fig.~\ref{fig:protAgeHist} by plotting a histogram of system ages from Fig.~\ref{fig:protDist} for single or primary stars in binaries with P$_{rot} < 20$ d.}

@@

Tidal torques pose a fundamental problem for inferring ages of stars via \xxx{gyrochronology}. Regardless of the choice of equilibrium tidal model or magnetic braking model, stellar binaries readily tidally-lock, or at least strongly tidally-interact, across a wide range of P$_{orb}$ and primary star masses, decoupling P$_{rot}$ from age. For example, if one observed a rapidly rotating star with P$_{rot} \lsim 20$ d, gyrochronology models would predict ages \xxx{$\lsim 1.5$ Gyr. If the star is actually an unresolved binary, as could be the case for many \kepler rapid rotators \citep{Simonian2018}, it would likely be tidally-locked, decoupling P$_{rot}$ from age, causing the predictions of gyrochronology models to fail}. This effect is most likely to manifest in rapid rotators (P$_{rot} < 20$ d), but persists across all P$_{rot}$ up to 100 d, producing a \xxx{contaminating signal, e.g. Fig.~\ref{fig:protDist} and Fig.~\ref{fig:gyro}}.

In general, it is difficult to accurately determine if a source is single star or a stellar binary via \xxx{longterm photometric monitoring, e.g. via \kepler or \textit{TESS},} as only a small fraction of stars in binaries will occult one another. Observations of the binarity of field stars by \citet{Raghavan2010} and \citet{Duchene2013} indicate that roughly half of stars are in stellar binaries, with $10\%$ of these binaries having P$_{orb} \lsim 100$ d, suggesting that unless one accounts for binarity, stellar binaries will produce a contaminating signal in any study of stellar rotation periods and any ages inferred via \xxx{gyrochronology} are potentially subject to systematic errors.  Moreover, this problem could be more significant as \citet{Simonian2018} found that most rapid rotators with P$_{rot} \leq 7.5$ d in the \kepler field \xxx{are consistent with tidally-synchronized photometric binaries, suggesting that binary contamination in P$_{rot}$ studies could be widespread}. We caution that any application of \xxx{gyrochronology} methods to predict ages for stars, especially those with P$_{rot} \lsim 20$ d, should rule out or account for stellar binarity, or otherwise risk deriving systematically incorrect ages.

@@

The torques due to tides and magnetic braking balance when $ \frac{dJ}{dt}|_{tides} +  \frac{dJ}{dt}|_{MB} = 0$, 
\begin{equation} \label{eqn:appendixBalance}
    \frac{C_{tides} k_2 \tau}{\mathrm{P}_{orb}^5}\left( 1 - \frac{\mathrm{P}_{orb}}{\mathrm{P}_{rot}} \right) -C_{MB} \left( \frac{P_{rot,\odot}}{P_{rot}} \right)^3 = 0.
\end{equation}
By specifying P$_{orb}$ and $k_2 \tau$, we can numerically solve Eqn.~(\ref{eqn:appendixBalance}) for the P$_{rot}$ at which torques due to magnetic braking and tides balance, often producing subsynchronous rotation as seen in Fig.~\ref{fig:analyticBalance} and our simulations in $\S$~\ref{sec:eq}.
